---
title: "Forest Impact Simulator"
author: "Forest Carbon Analysis Tool"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(ggplot2)
library(jsonlite)
library(readr)
library(dplyr)
library(plotly)
```

# Forest Impact Simulator

R companion to the [Forest Impact Simulator web app](https://forest-impact-simulator.vercel.app/).
Calculates carbon sequestration or loss, biodiversity, resilience, water retention, and air quality
using the same growth curves and environmental modifiers as the TypeScript implementation.

## 1. Constants (aligned with web app)

```{r constants}
# Planting-mode growth factors by simulation year (from src/utils/constants.ts)
TREE_GROWTH_FACTORS <- list(
  year_1 = 0.05,
  year_2 = 0.15,
  year_3 = 0.30,
  year_4 = 0.50,
  year_5 = 0.70,
  year_6_to_10 = 0.80,
  year_11_to_20 = 0.90,
  year_20_plus = 1.00
)

# Clear-cutting age-based growth factors
TREE_AGE_GROWTH_FACTORS <- list(
  age_1 = 0.05,
  age_2 = 0.15,
  age_3 = 0.30,
  age_4 = 0.50,
  age_5 = 0.70,
  age_6 = 0.85,
  age_7_to_20 = 0.95,
  age_21_to_50 = 0.90,
  age_50_plus = 0.85
)

SOIL_CARBON_MODIFIER <- 0.1          # kg CO2/year per g/kg soil carbon
PRECIP_TO_RESILIENCE <- 0.001
BIODIVERSITY_TIME_BONUS <- 0.05
RESILIENCE_TIME_BONUS <- 0.03
WATER_ANNUAL_IMPROVEMENT <- 0.3
WATER_ANNUAL_DEGRADATION <- 0.5
AIR_ANNUAL_IMPROVEMENT <- 0.7
AIR_ANNUAL_DEGRADATION <- 1.0
MAX_WATER <- 95
MAX_AIR <- 95
MAX_AIR_DEGRADATION <- -80
CAR_EMISSIONS_PER_YEAR <- 4600       # kg CO2
FLIGHT_NY_LONDON <- 986              # kg CO2
HOUSEHOLD_ELECTRICITY_PER_YEAR <- 7500

get_planting_growth_factor <- function(year) {
  if (year <= 1) return(TREE_GROWTH_FACTORS$year_1)
  if (year == 2) return(TREE_GROWTH_FACTORS$year_2)
  if (year == 3) return(TREE_GROWTH_FACTORS$year_3)
  if (year == 4) return(TREE_GROWTH_FACTORS$year_4)
  if (year == 5) return(TREE_GROWTH_FACTORS$year_5)
  if (year <= 10) return(TREE_GROWTH_FACTORS$year_6_to_10)
  if (year <= 20) return(TREE_GROWTH_FACTORS$year_11_to_20)
  TREE_GROWTH_FACTORS$year_20_plus
}

get_age_growth_factor <- function(age) {
  if (age <= 1) return(TREE_AGE_GROWTH_FACTORS$age_1)
  if (age <= 2) return(TREE_AGE_GROWTH_FACTORS$age_2)
  if (age <= 3) return(TREE_AGE_GROWTH_FACTORS$age_3)
  if (age <= 4) return(TREE_AGE_GROWTH_FACTORS$age_4)
  if (age <= 5) return(TREE_AGE_GROWTH_FACTORS$age_5)
  if (age <= 6) return(TREE_AGE_GROWTH_FACTORS$age_6)
  if (age <= 20) return(TREE_AGE_GROWTH_FACTORS$age_7_to_20)
  if (age <= 50) return(TREE_AGE_GROWTH_FACTORS$age_21_to_50)
  TREE_AGE_GROWTH_FACTORS$age_50_plus
}

# Cumulative carbon for planting (mature annual rate × growth curve)
calculate_cumulative_carbon <- function(mature_annual_rate, years) {
  total <- 0
  for (year in seq_len(years)) {
    total <- total + mature_annual_rate * get_planting_growth_factor(year)
  }
  total
}

# Clear-cutting: lifetime stored carbon + lost future sequestration
calculate_clear_cutting_carbon <- function(mature_annual_rate, tree_age, simulation_years) {
  immediate <- 0
  for (year in seq_len(tree_age)) {
    immediate <- immediate + mature_annual_rate * get_age_growth_factor(year)
  }
  lost_future <- 0
  for (year in seq_len(simulation_years)) {
    lost_future <- lost_future + mature_annual_rate * get_age_growth_factor(tree_age + year)
  }
  list(immediate = immediate, lost_future = lost_future, total = immediate + lost_future)
}

split_semicolon <- function(value) {
  if (length(value) == 0 || is.null(value) || (length(value) == 1 && is.na(value))) {
    return(character(0))
  }
  text <- as.character(value)
  if (!nzchar(text)) return(character(0))
  trimws(unlist(strsplit(text, ";", fixed = TRUE)))
}
```

## 2. Data Input

### Supported formats

1. **Web app CSV** (recommended) — single header + one data row from
   [forest-impact-simulator.vercel.app](https://forest-impact-simulator.vercel.app/) export
2. **Plot CSV** — one row per land plot (`land_id`, `area_ha`, `tree_type`, `carbon_per_tree`, …)
3. **Manual input** — edit the data frames below

```{r data-input, echo=FALSE, message=FALSE}
detect_csv_format <- function(file_path) {
  headers <- tolower(names(read.csv(file_path, nrows = 0, check.names = FALSE)))
  if ("timestamp" %in% headers && "tree_names" %in% headers) {
    return("webapp")
  }
  if ("land_id" %in% headers || "plot_id" %in% headers) {
    return("plot")
  }
  "unknown"
}

parse_webapp_csv <- function(file_path) {
  row <- read.csv(file_path, stringsAsFactors = FALSE, check.names = FALSE)[1, , drop = FALSE]

  tree_names <- split_semicolon(row$tree_names)
  tree_scientific <- split_semicolon(row$tree_scientific_names)
  tree_rates <- as.numeric(split_semicolon(row$tree_carbon_rates_kg_co2_year))
  tree_pcts <- as.numeric(split_semicolon(row$tree_percentages))

  n_trees <- max(length(tree_names), 1)
  if (length(tree_scientific) < n_trees) tree_scientific <- rep(NA_character_, n_trees)
  if (length(tree_rates) < n_trees) tree_rates <- rep(NA_real_, n_trees)
  # Web exports sometimes store 0 for a single species; treat missing/zero as equal split
  if (length(tree_pcts) < n_trees || all(is.na(tree_pcts)) || sum(tree_pcts, na.rm = TRUE) <= 0) {
    tree_pcts <- rep(100 / n_trees, n_trees)
  }

  tree_data <- data.frame(
    name = tree_names,
    scientific_name = tree_scientific[seq_len(n_trees)],
    carbon_per_tree = tree_rates[seq_len(n_trees)],
    percentage = tree_pcts[seq_len(n_trees)],
    stringsAsFactors = FALSE
  )

  list(
    format = "webapp",
    row = row,
    tree_data = tree_data,
    simulation_years = as.numeric(row$simulation_years),
    latitude = as.numeric(row$latitude),
    longitude = as.numeric(row$longitude),
    area_ha = as.numeric(row$area_hectares),
    total_trees = as.numeric(row$total_trees),
    spacing = as.numeric(row$spacing_meters),
    density = as.numeric(row$density_trees_hectare),
    soil_carbon = suppressWarnings(as.numeric(row$soil_carbon_g_kg)),
    soil_ph = suppressWarnings(as.numeric(row$soil_ph)),
    temperature = suppressWarnings(as.numeric(row$temperature_c)),
    precipitation = suppressWarnings(as.numeric(row$precipitation_mm)),
    # Pre-computed impact results from the web app (use when present)
    annual_carbon = suppressWarnings(as.numeric(row$annual_carbon_sequestration_kg_co2_year)),
    total_carbon = suppressWarnings(as.numeric(row$total_carbon_kg_co2)),
    biodiversity = suppressWarnings(as.numeric(row$biodiversity_impact)),
    resilience = suppressWarnings(as.numeric(row$forest_resilience)),
    water_retention = suppressWarnings(as.numeric(row$water_retention_percent)),
    air_quality = suppressWarnings(as.numeric(row$air_quality_improvement_percent)),
    average_biodiversity = suppressWarnings(as.numeric(row$average_biodiversity)),
    average_resilience = suppressWarnings(as.numeric(row$average_resilience))
  )
}

parse_plot_csv <- function(file_path) {
  df <- read.csv(file_path, stringsAsFactors = FALSE, check.names = FALSE)
  if (!"land_id" %in% names(df) && "plot_id" %in% names(df)) {
    df$land_id <- df$plot_id
  }
  if (!"carbon_per_tree" %in% names(df) && "carbon_per_ha" %in% names(df)) {
    # Legacy column name: treat as per-tree mature rate (kg CO2/year/tree)
    df$carbon_per_tree <- df$carbon_per_ha
  }
  list(format = "plot", land_data = df)
}

webapp_file <- "sample-forest-planting-data.csv"
plot_file <- "sample-forest-plot-data.csv"
has_csv_data <- FALSE
csv_format <- "none"
parsed <- NULL

if (file.exists(webapp_file)) {
  fmt <- detect_csv_format(webapp_file)
  if (fmt == "webapp") {
    cat("**Web app CSV found!** Loading comprehensive export...\n\n")
    parsed <- parse_webapp_csv(webapp_file)
    has_csv_data <- TRUE
    csv_format <- "webapp"
    cat("**Trees:**\n")
    print(parsed$tree_data)
    cat("\n**Key fields:**\n")
    cat(sprintf("- Years: %s | Area: %s ha | Trees: %s\n",
                parsed$simulation_years, parsed$area_ha, parsed$total_trees))
    cat(sprintf("- Location: %s, %s\n", parsed$latitude, parsed$longitude))
    cat(sprintf("- Pre-computed total carbon: %s kg CO2\n",
                format(parsed$total_carbon, big.mark = ",")))
  } else if (fmt == "plot") {
    cat("**Plot CSV found** in sample-forest-planting-data.csv\n\n")
    parsed <- parse_plot_csv(webapp_file)
    has_csv_data <- TRUE
    csv_format <- "plot"
    print(parsed$land_data)
  } else {
    cat("**Unknown CSV format** in sample-forest-planting-data.csv. Using manual input.\n\n")
  }
} else if (file.exists(plot_file)) {
  cat("**Plot CSV found!** Loading plot-level data...\n\n")
  parsed <- parse_plot_csv(plot_file)
  has_csv_data <- TRUE
  csv_format <- "plot"
  print(parsed$land_data)
} else {
  cat("**No CSV data files found.** Using manual input below.\n\n")
}
```

### Option B: Manual Data Input

Used when no CSV is present, or as a template for custom runs.

```{r manual-input, echo=FALSE}
# Manual plot-level input (carbon_per_tree = mature kg CO2/year/tree)
manual_land_data <- data.frame(
  land_id = c("Plot_001", "Plot_002", "Plot_003"),
  area_ha = c(0.1, 0.15, 0.2),
  tree_type = c("Oak", "Pine", "Mixed"),
  carbon_per_tree = c(22, 18.5, 20),
  trees_per_ha = c(625, 625, 625),
  biodiversity_score = c(4.5, 4.2, 4.8),
  resilience_score = c(4.3, 4.0, 4.6),
  stringsAsFactors = FALSE
)

manual_location <- data.frame(
  latitude = 50.56956736416948,
  longitude = 28.159344338632376,
  simulation_years = 50,
  tree_age = 20   # used only for clear-cutting
)

manual_env <- data.frame(
  soil_carbon = NA_real_,
  soil_ph = NA_real_,
  temperature = 15,
  precipitation = NA_real_
)

if (!has_csv_data) {
  cat("**Manual Land Plot Data:**\n")
  print(manual_land_data)
  cat("\n**Location / Years:**\n")
  print(manual_location)
}
```

## 3. Mode Selection

```{r mode-selection, echo=FALSE}
mode <- "planting"   # or "clear-cutting"
tree_age <- 20       # average age of existing trees (clear-cutting only)

cat(paste("**Selected Mode:**", toupper(mode), "\n\n"))
if (mode == "planting") {
  cat("Planting mode: cumulative carbon with realistic growth curve.\n")
} else {
  cat(sprintf(
    "Clear-cutting mode: immediate release (age %s) + lost future sequestration.\n",
    tree_age
  ))
}
```

## 4. Calculations

```{r calculations, echo=TRUE}
if (has_csv_data && csv_format == "webapp") {
  simulation_years <- parsed$simulation_years
  latitude <- parsed$latitude
  longitude <- parsed$longitude
  area_ha <- parsed$area_ha
  total_trees <- parsed$total_trees
  spacing <- ifelse(is.na(parsed$spacing), 4, parsed$spacing)
  density <- ifelse(is.na(parsed$density), 625, parsed$density)

  soil_carbon <- parsed$soil_carbon
  soil_ph <- parsed$soil_ph
  temperature <- parsed$temperature
  precipitation <- parsed$precipitation

  # Weighted mature rate per tree (kg CO2/year)
  weights <- parsed$tree_data$percentage / 100
  if (any(is.na(weights)) || sum(weights) == 0) {
    weights <- rep(1 / nrow(parsed$tree_data), nrow(parsed$tree_data))
  }
  mature_rate_per_tree <- sum(parsed$tree_data$carbon_per_tree * weights, na.rm = TRUE)

  # Soil carbon modifier (matches web app)
  if (!is.na(soil_carbon)) {
    mature_rate_per_tree <- mature_rate_per_tree + soil_carbon * SOIL_CARBON_MODIFIER
  }

  mature_annual_total <- mature_rate_per_tree * total_trees

  if (mode == "planting") {
    recalculated_total <- calculate_cumulative_carbon(mature_annual_total, simulation_years)
    immediate_release <- NA_real_
    lost_future <- NA_real_
  } else {
    cc <- calculate_clear_cutting_carbon(mature_annual_total, tree_age, simulation_years)
    recalculated_total <- -cc$total
    immediate_release <- cc$immediate
    lost_future <- cc$lost_future
  }

  # Prefer web-app precomputed totals for planting when available;
  # always recalculate for clear-cutting (export is planting-oriented).
  if (mode == "planting" && !is.na(parsed$total_carbon)) {
    total_carbon <- parsed$total_carbon
    annual_carbon <- ifelse(
      !is.na(parsed$annual_carbon),
      parsed$annual_carbon,
      total_carbon / simulation_years
    )
    used_source <- "web app export"
  } else {
    total_carbon <- recalculated_total
    annual_carbon <- total_carbon / simulation_years
    used_source <- "recalculated growth curve"
  }

  biodiversity <- ifelse(!is.na(parsed$biodiversity), parsed$biodiversity, 4.5)
  resilience <- ifelse(!is.na(parsed$resilience), parsed$resilience, 4.5)
  if (!is.na(precipitation) && is.na(parsed$resilience)) {
    resilience <- min(5, resilience + precipitation * PRECIP_TO_RESILIENCE)
  }
  water_retention <- ifelse(!is.na(parsed$water_retention), parsed$water_retention, 85)
  air_quality <- ifelse(!is.na(parsed$air_quality), parsed$air_quality, 80)
  average_biodiversity <- ifelse(
    !is.na(parsed$average_biodiversity), parsed$average_biodiversity, biodiversity
  )
  average_resilience <- ifelse(
    !is.na(parsed$average_resilience), parsed$average_resilience, resilience
  )

  land_data <- data.frame(
    land_id = paste0("Species_", seq_len(nrow(parsed$tree_data))),
    tree_type = parsed$tree_data$name,
    scientific_name = parsed$tree_data$scientific_name,
    percentage = parsed$tree_data$percentage,
    carbon_per_tree = parsed$tree_data$carbon_per_tree,
    trees = round(total_trees * weights),
    area_ha = area_ha * weights,
    stringsAsFactors = FALSE
  )
  land_data$co2_impact_kg <- if (mode == "planting") {
    vapply(seq_len(nrow(land_data)), function(i) {
      calculate_cumulative_carbon(
        land_data$carbon_per_tree[i] * land_data$trees[i],
        simulation_years
      )
    }, numeric(1))
  } else {
    vapply(seq_len(nrow(land_data)), function(i) {
      -calculate_clear_cutting_carbon(
        land_data$carbon_per_tree[i] * land_data$trees[i],
        tree_age,
        simulation_years
      )$total
    }, numeric(1))
  }

} else {
  # Plot CSV or manual input
  if (has_csv_data && csv_format == "plot") {
    land_data <- parsed$land_data
    simulation_years <- if ("simulation_years" %in% names(land_data)) {
      land_data$simulation_years[1]
    } else {
      manual_location$simulation_years
    }
    latitude <- if ("latitude" %in% names(land_data)) land_data$latitude[1] else manual_location$latitude
    longitude <- if ("longitude" %in% names(land_data)) land_data$longitude[1] else manual_location$longitude
    soil_carbon <- if ("soil_carbon" %in% names(land_data)) land_data$soil_carbon[1] else NA_real_
    soil_ph <- if ("soil_ph" %in% names(land_data)) land_data$soil_ph[1] else NA_real_
    temperature <- if ("temperature" %in% names(land_data)) land_data$temperature[1] else 15
    precipitation <- if ("precipitation" %in% names(land_data)) land_data$precipitation[1] else NA_real_
  } else {
    land_data <- manual_land_data
    simulation_years <- manual_location$simulation_years
    latitude <- manual_location$latitude
    longitude <- manual_location$longitude
    tree_age <- manual_location$tree_age
    soil_carbon <- manual_env$soil_carbon
    soil_ph <- manual_env$soil_ph
    temperature <- manual_env$temperature
    precipitation <- manual_env$precipitation
  }

  if (!"trees_per_ha" %in% names(land_data)) land_data$trees_per_ha <- 625
  if (!"carbon_per_tree" %in% names(land_data)) {
    stop("Plot data needs carbon_per_tree (mature kg CO2/year/tree).")
  }
  land_data$trees <- round(land_data$area_ha * land_data$trees_per_ha)

  land_data$co2_impact_kg <- if (mode == "planting") {
    vapply(seq_len(nrow(land_data)), function(i) {
      rate <- land_data$carbon_per_tree[i]
      if (!is.na(soil_carbon)) rate <- rate + soil_carbon * SOIL_CARBON_MODIFIER
      calculate_cumulative_carbon(rate * land_data$trees[i], simulation_years)
    }, numeric(1))
  } else {
    vapply(seq_len(nrow(land_data)), function(i) {
      rate <- land_data$carbon_per_tree[i]
      if (!is.na(soil_carbon)) rate <- rate + soil_carbon * SOIL_CARBON_MODIFIER
      -calculate_clear_cutting_carbon(
        rate * land_data$trees[i], tree_age, simulation_years
      )$total
    }, numeric(1))
  }

  total_carbon <- sum(land_data$co2_impact_kg)
  annual_carbon <- total_carbon / simulation_years
  recalculated_total <- total_carbon
  immediate_release <- NA_real_
  lost_future <- NA_real_
  used_source <- "recalculated growth curve"

  area_ha <- sum(land_data$area_ha)
  total_trees <- sum(land_data$trees)
  spacing <- 4
  density <- 625

  bio_col <- if ("biodiversity_score" %in% names(land_data)) {
    land_data$biodiversity_score
  } else {
    rep(4.0, nrow(land_data))
  }
  res_col <- if ("resilience_score" %in% names(land_data)) {
    land_data$resilience_score
  } else {
    rep(4.0, nrow(land_data))
  }

  time_sign <- if (mode == "planting") 1 else -1
  biodiversity <- min(5, max(0, mean(bio_col) + time_sign * simulation_years * BIODIVERSITY_TIME_BONUS))
  resilience <- mean(res_col)
  if (!is.na(precipitation)) {
    resilience <- resilience + precipitation * PRECIP_TO_RESILIENCE
  }
  resilience <- min(5, max(0, resilience + time_sign * simulation_years * RESILIENCE_TIME_BONUS))

  abs_lat <- abs(latitude)
  water_base <- if (abs_lat < 30) 85 else if (abs_lat < 60) 75 else 70
  if (mode == "planting") {
    water_retention <- min(MAX_WATER, water_base + simulation_years * WATER_ANNUAL_IMPROVEMENT)
    air_base <- if (abs_lat < 30) 70 else if (abs_lat < 60) 60 else 50
    air_quality <- min(MAX_AIR, air_base + simulation_years * AIR_ANNUAL_IMPROVEMENT)
  } else {
    water_retention <- max(0, water_base - simulation_years * WATER_ANNUAL_DEGRADATION)
    immediate_air <- min(30, log10(max(total_trees, 10)) * 5)
    air_quality <- max(MAX_AIR_DEGRADATION, -(immediate_air + simulation_years * AIR_ANNUAL_DEGRADATION))
  }

  average_biodiversity <- biodiversity
  average_resilience <- resilience
}

impact_type <- if (mode == "planting") "Carbon Sequestration" else "Carbon Emissions"
comparisons <- list(
  car_years = abs(total_carbon) / CAR_EMISSIONS_PER_YEAR,
  flights = abs(total_carbon) / FLIGHT_NY_LONDON,
  household_years = abs(total_carbon) / HOUSEHOLD_ELECTRICITY_PER_YEAR
)

cat("## Calculation Results\n\n")
cat(sprintf("**Source:** %s\n\n", used_source))
cat("**Plot / species breakdown:**\n")
print(land_data)

cat("\n**Summary:**\n")
cat(sprintf("- Total %s: %s kg CO2\n", impact_type, format(round(total_carbon, 1), big.mark = ",")))
cat(sprintf("- Period-average annual: %s kg CO2/year\n", format(round(annual_carbon, 1), big.mark = ",")))
if (mode == "clear-cutting" && !is.na(immediate_release)) {
  cat(sprintf("- Immediate release: %s kg CO2\n", format(round(immediate_release, 1), big.mark = ",")))
  cat(sprintf("- Lost future sequestration: %s kg CO2\n", format(round(lost_future, 1), big.mark = ",")))
}
cat(sprintf("- Area: %s ha | Trees: %s | Years: %s\n",
            round(area_ha, 2), format(total_trees, big.mark = ","), simulation_years))
cat(sprintf("- Biodiversity: %.1f/5 | Resilience: %.1f/5\n", biodiversity, resilience))
cat(sprintf("- Water retention: %.0f%% | Air quality: %.0f%%\n", water_retention, air_quality))
cat("\n**Real-world comparisons (absolute CO2):**\n")
cat(sprintf("- ≈ %.0f years of average car emissions\n", comparisons$car_years))
cat(sprintf("- ≈ %.0f NY–London round-trip flights\n", comparisons$flights))
cat(sprintf("- ≈ %.0f years of US household electricity\n", comparisons$household_years))

if (has_csv_data && csv_format == "webapp" && mode == "planting") {
  cat(sprintf(
    "\n**Growth-curve recalculation check:** %s kg CO2 (web app export: %s)\n",
    format(round(recalculated_total, 1), big.mark = ","),
    format(round(parsed$total_carbon, 1), big.mark = ",")
  ))
}
```

## 5. Visualizations

### CO2 Impact by Species / Plot

```{r co2-bar-chart, fig.width=10, fig.height=6}
y_col <- if ("co2_impact_kg" %in% names(land_data)) "co2_impact_kg" else "co2_impact_tons"
x_col <- if ("tree_type" %in% names(land_data)) "tree_type" else "land_id"

plot_df <- land_data
plot_df$label_x <- plot_df[[x_col]]
plot_df$co2 <- plot_df[[y_col]]

co2_plot <- ggplot(plot_df, aes(x = label_x, y = co2, fill = tree_type)) +
  geom_bar(stat = "identity", alpha = 0.85) +
  geom_text(aes(label = paste(format(round(co2, 0), big.mark = ","), "kg")),
            vjust = -0.4, size = 3.2) +
  labs(
    title = paste("CO2 Impact —", impact_type),
    subtitle = paste("Total:", format(round(total_carbon, 1), big.mark = ","),
                     "kg CO2 over", simulation_years, "years"),
    x = if (x_col == "tree_type") "Tree Type" else "Land Plot",
    y = "CO2 (kg)",
    fill = "Tree Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(angle = 30, hjust = 1),
    legend.position = "bottom"
  ) +
  scale_fill_brewer(type = "qual", palette = "Set2")

print(co2_plot)
```

### Environmental Metrics

```{r metrics-comparison, fig.width=9, fig.height=5}
metrics_df <- data.frame(
  metric = c("Biodiversity", "Resilience", "Water retention /20", "Air quality /20"),
  score = c(
    biodiversity,
    resilience,
    water_retention / 20,
    abs(air_quality) / 20
  )
)

metrics_plot <- ggplot(metrics_df, aes(x = metric, y = score, fill = metric)) +
  geom_bar(stat = "identity", alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = round(score, 2)), vjust = -0.4, size = 3.5) +
  labs(
    title = "Environmental Metrics",
    subtitle = "Biodiversity & resilience on 0–5 scale; water/air scaled /20 for comparison",
    x = NULL,
    y = "Score"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold")) +
  ylim(0, 5.5) +
  scale_fill_brewer(type = "qual", palette = "Set1")

print(metrics_plot)
```

### Yearly Carbon Accumulation (growth curve)

```{r yearly-projection, fig.width=12, fig.height=6}
if (mode == "planting" && simulation_years > 1) {
  # Use mature annual rate implied by period-average and growth factors
  mature_annual <- if (has_csv_data && csv_format == "webapp") {
    mature_annual_total
  } else {
    # Back out approximate mature rate from total trees × mean per-tree rate
    mean(land_data$carbon_per_tree, na.rm = TRUE) * total_trees
  }

  yearly <- data.frame(year = seq_len(simulation_years))
  yearly$annual <- sapply(yearly$year, function(y) {
    mature_annual * get_planting_growth_factor(y)
  })
  yearly$cumulative <- cumsum(yearly$annual)

  yearly_plot <- ggplot(yearly, aes(x = year, y = cumulative)) +
    geom_line(color = "#2E8B57", linewidth = 1.2) +
    geom_area(fill = "#2E8B57", alpha = 0.2) +
    labs(
      title = "Cumulative Carbon Sequestration (Growth Curve)",
      subtitle = "Year 1 ≈ 5% of mature rate → Year 20+ ≈ 100%",
      x = "Year",
      y = "Cumulative Carbon (kg CO2)"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(size = 16, face = "bold"))

  print(yearly_plot)
}
```

## 6. Export

```{r save-outputs}
if (!dir.exists("output")) dir.create("output")

output_csv <- "output/processed_forest_data.csv"
write.csv(land_data, output_csv, row.names = FALSE)
cat(paste("Processed data saved to:", output_csv, "\n"))

timestamp <- format(Sys.time(), "%Y-%m-%dT%H:%M:%S.000Z")

# Web-app-compatible JSON (matches ExportData shape)
json_output <- list(
  metadata = list(
    timestamp = timestamp,
    simulatorVersion = "1.0.0",
    location = list(
      latitude = latitude,
      longitude = longitude,
      type = "Point"
    ),
    simulation = list(
      years = simulation_years,
      mode = mode,
      totalPlots = nrow(land_data)
    )
  ),
  environmentalData = list(
    soil = list(
      carbon = if (is.na(soil_carbon)) NULL else soil_carbon,
      pH = if (is.na(soil_ph)) NULL else soil_ph
    ),
    climate = list(
      temperature = if (is.na(temperature)) NULL else temperature,
      precipitation = if (is.na(precipitation)) NULL else precipitation
    )
  ),
  impactResults = list(
    carbonSequestration = annual_carbon,
    biodiversityImpact = biodiversity,
    forestResilience = resilience,
    waterRetention = water_retention,
    airQualityImprovement = air_quality,
    totalCarbon = total_carbon,
    averageBiodiversity = average_biodiversity,
    averageResilience = average_resilience
  ),
  plantingData = list(
    area = area_ha,
    totalTrees = total_trees,
    spacing = spacing,
    density = density,
    timeline = simulation_years,
    plots = land_data
  )
)

output_json <- "output/forest_impact_results.json"
write_json(json_output, output_json, pretty = TRUE, auto_unbox = TRUE, null = "null")
cat(paste("JSON results saved to:", output_json, "\n\n"))

cat("## Final Summary\n\n")
cat(sprintf("- Mode: %s\n", toupper(mode)))
cat(sprintf("- Total %s: %s kg CO2\n", impact_type, format(round(total_carbon, 1), big.mark = ",")))
cat(sprintf("- Years: %s | Area: %s ha | Trees: %s\n",
            simulation_years, round(area_ha, 2), format(total_trees, big.mark = ",")))
```

---

## Instructions

1. **CSV from web app**: Export CSV on the live site and save as `sample-forest-planting-data.csv` (or replace the included sample).
2. **Mode**: Set `mode` to `"planting"` or `"clear-cutting"`; set `tree_age` for clear-cutting.
3. **Run** all chunks in RStudio; outputs land in `output/`.
4. **Growth curve**: Planting carbon uses the documented year-by-year factors (5% → 100%), not linear `rate × years`.

## Notes

- Web app CSV columns match `src/utils/exportUtils.ts` in the TypeScript repo.
- Tree carbon rates are **kg CO₂/year per mature tree**, scaled by tree count — not per hectare.
- Clear-cutting totals = lifetime stored carbon + lost future sequestration.
- Constants mirror [Forest-Impact-Simulator](https://github.com/karimogit/Forest-Impact-Simulator) `src/utils/constants.ts`.

*Forest Impact Simulator R Notebook — aligned with web app v1.0.0*
